W---
title: "Design Declaration"
subtitle: "Punish or Reconcile? Target Framing, Performance Cues, and Authoritarian Legitimation through Repression"
author:
    - Daniel Kuhlen
    - Giovanna Lapresa
    - Jorge Zavala
date: today
format:
  html:
    embed-resources: true
bibliography: ../../05_bibliography/bib.bib
---

```{r}
#| include: false

# packages
pacman::p_load(tidyverse,
               rio,
               DeclareDesign,
               estimatr,
               scales)
```

# Theory

This experiment examines how citizens update their support for repressive government policies based on three informational dimensions: the nature of the justice policy (punitive vs. restorative), the framing of the targeted group (as a threat or not), and performance information about the policy (positive, negative, or absent).

We build on @lachapelle2022's triadic model of repression, which theorizes that bystander citizens evaluate repressive policies through a legitimacy lens, and extend it with a performance dimension following @przeworski2023.

## Mechanism: Implicit Performance Beliefs

Our central hypothesis is that citizens support punitive justice policies partly because they *implicitly assume* these policies are more effective---especially when the targeted group is framed as a threat. The performance dimension in our design allows us to test this mechanism.

Following @acharya2018, we treat performance information as a **manipulated mediator** for the justice frame effect. When no performance information is provided (the "natural-mediator" condition), respondents rely on their prior beliefs about policy effectiveness---which we hypothesize are biased toward assuming punitive policies work better. When performance information *is* provided (the "manipulated-mediator" condition), these implicit beliefs are overridden by the experimental cue, and the justice frame effect should change.

The **eliminated effect** [@acharya2018]---the difference between the justice frame effect with and without performance information---captures how much of the punitive justice premium is explained by implicit performance beliefs (through either indirect effects or causal interactions). A large eliminated effect would indicate that performance assumptions are a key mechanism driving support for punitive repression.

# Design

We implement a **2 x 2 x 3 conjoint-style vignette experiment** with repeated measures. Each respondent reads a series of short vignettes describing a government policy and is asked to rate their support for the policy on a 0--10 scale after each one.

## Factors

The three experimental factors and their levels are:

| Factor | Levels | Description |
|--------|--------|-------------|
| **Justice Policy** | Punitive, Restorative | Whether the policy is framed as punitive or restorative |
| **Threat Framing** | Threat, No Threat | Whether the targeted group is described as a threat |
| **Performance Information** | None, Positive, Negative | Whether and what kind of policy performance information is provided |

This factorial crossing produces **12 unique vignette types** (2 x 2 x 3 = 12).

## Within-Subjects Design

Each respondent is randomly assigned to see **K = 5** vignette scenarios. After reading each vignette, the respondent answers the same outcome battery (support for repressive policy on a 0--10 scale). This within-subjects repeated-measures design dramatically increases the effective analytical sample: with N respondents, we observe N x K = N x 5 respondent-vignette observations.

## Randomization Strategy

Following @hainmueller2014, the three attribute levels are **independently randomized** for each profile (i.e., the level of one attribute does not depend on the levels of the others). Each respondent sees **K = 5 distinct vignette types** sampled without replacement from the 12 possible combinations, ensuring variety across tasks. The key identification assumptions for AMCEs are:

1. **No carryover effects**: a respondent's evaluation of one vignette is not affected by previously seen vignettes
2. **No profile-order effects**: the position of a vignette in the sequence does not affect evaluations
3. **Independent attribute randomization**: attribute levels are uncorrelated within each profile

Per @humphreys2026, we are explicit that our estimands are **effects of information** (the vignette content) on stated policy support, not effects on underlying beliefs or latent preferences. We also note the risk of cross-attribute belief updating: seeing a "punitive" justice frame may cause respondents to infer threat, which we address through the fully randomized design that breaks any structural correlation between attributes.

## Causal Mechanisms via Manipulated Mediator

Following @acharya2018, the 2 x 2 x 3 design can be read as a conjoint experiment with an embedded manipulated-mediator structure:

- **Natural-mediator arm** (performance = "none"): Only the justice and threat frames are varied. Respondents rely on their *implicit beliefs* about policy effectiveness. In this arm, we estimate the **total effect** of the justice frame, which bundles together direct preferences and any indirect effect flowing through assumed effectiveness.
- **Manipulated-mediator arms** (performance = "positive" or "negative"): Performance information is provided, fixing the respondent's information about policy effectiveness at a particular value. In these arms, we estimate the **controlled direct effect** (ACDE) of the justice frame---the part of the effect that persists even when effectiveness beliefs are experimentally overridden.

The **eliminated effect** is the difference: $\Delta = \text{ATE}_{\text{justice}} - \text{ACDE}_{\text{justice}}$. A positive eliminated effect indicates that implicit performance beliefs explain part of the punitive justice premium, either through mediation (punitive → assumed effective → support) or interaction (the justice effect depends on effectiveness context). Following @acharya2018, this design avoids the strong sequential ignorability assumption of traditional mediation analysis by *randomizing* the mediator rather than merely observing it.

## Design Declaration

The **model** assumes additive main effects for the three factors on support for repressive policy (0--10 scale), plus an interaction between the justice frame and performance information provision---reflecting the hypothesis that providing performance information changes the justice frame effect. The data-generating process includes a respondent-level random intercept and task-level noise, producing within-respondent correlation (ICC $\approx$ 0.8) consistent with repeated attitudinal measures. Baseline parameters (population mean $\mu$ = 4, standard deviation $\sigma$ = 2) are placeholders.

The **inquiry** consists of four AMCEs, plus the eliminated effect following @acharya2018: the difference between the justice frame AMCE in the natural-mediator arm (no performance info) and in the manipulated-mediator arms (performance info provided). This captures how much of the punitive justice premium is explained by implicit performance beliefs.

The **data strategy** implements full independent randomization of attribute levels for each respondent-task observation, following the conjoint design described above.

The **answer strategy** uses OLS with CR2 cluster-robust standard errors at the respondent level, following @hainmueller2014's recommendation for conjoint designs with multiple tasks per respondent. The eliminated effect is estimated via a justice $\times$ performance-info interaction term.

```{r}
#| label: design-declaration

# set seed
set.seed(030)

# wrap the design in a function for parameter variation
declare_conjoint_design <- function(

  # number of survey respondents
  N              = 2000,
  # number of vignette tasks per respondent
  K              = 5,
  # baseline mean support for repressive policy (0-10 scale)
  mu             = 4,
  # population standard deviation (respondent-level)
  sigma          = 2,
  # AMCE of punitive (vs. restorative) justice frame
  tau_justice    = 0.8,
  # AMCE of threat (vs. no threat) framing
  tau_threat     = 1.2,
  # AMCE of positive performance info (vs. no info) - expected to reduce support
  tau_perf_pos   = -0.5,
  # AMCE of negative performance info (vs. no info) - expected to increase support
  tau_perf_neg   = 0.5,
  # interaction: how much the justice effect changes when performance info is provided
  # negative = providing performance info reduces the punitive premium (as hypothesized)
  # following Acharya, Blackwell & Sen (2018): eliminated effect = -tau_interact
  tau_interact   = -0.4
) {

  # total number of respondent-task observations
  N_obs <- N * K

  # m: model ----------------------------------------------------------------
  # each row is a respondent-task pair

  model <- declare_model(
    N = N_obs,
    respondent_id = rep(1:(N/K), each = K),
    task = rep(1:K, times = N/K),
    # respondent-level random intercept (constant within respondent)
    u_i = rep(rnorm(N/K, 0, sigma), each = K),
    # task-level noise (varies across tasks within respondent)
    e_ij = rnorm(N, 0, sigma * 0.5)
  )

  # d: data strategy --------------------------------------------------------
  # fully randomize each attribute independently for each respondent-task
  # following Hainmueller, Hopkins & Yamamoto (2014)

  assign_justice <- declare_assignment(
    Z_justice = complete_ra(N = N_obs, prob = 0.5)
  )

  assign_threat <- declare_assignment(
    Z_threat = complete_ra(N = N_obs, prob = 0.5)
  )

  # performance has 3 levels: "none" (baseline), "positive", "negative"
  assign_performance <- declare_assignment(
    Z_performance = complete_ra(
      N = N_obs,
      conditions = c("none", "positive", "negative")
    )
  )

  # measurement: construct outcome with explicit dummy variables
  # includes justice x performance-info interaction (Acharya et al. 2018)
  measure_Y <- declare_measurement(
    perf_positive = as.numeric(Z_performance == "positive"),
    perf_negative = as.numeric(Z_performance == "negative"),
    has_perf_info = as.numeric(Z_performance != "none"),
    Y = pmin(10, pmax(0,
      mu +
      tau_justice   * Z_justice +
      tau_threat    * Z_threat +
      tau_perf_pos  * perf_positive +
      tau_perf_neg  * perf_negative +
      tau_interact  * Z_justice * has_perf_info +
      u_i + e_ij
    ))
  )

  # i: inquiry --------------------------------------------------------------
  # AMCEs: the average effect of switching each attribute from baseline

  inquiry_justice <- declare_inquiry(
    AMCE_Justice = tau_justice,
    label = "AMCE: Punitive vs. Restorative"
  )

  inquiry_threat <- declare_inquiry(
    AMCE_Threat = tau_threat,
    label = "AMCE: Threat vs. No Threat"
  )

  inquiry_perf_pos <- declare_inquiry(
    AMCE_Perf_Positive = tau_perf_pos,
    label = "AMCE: Positive Performance vs. None"
  )

  inquiry_perf_neg <- declare_inquiry(
    AMCE_Perf_Negative = tau_perf_neg,
    label = "AMCE: Negative Performance vs. None"
  )

  # justice x performance-info interaction (Acharya et al. 2018)
  # this is the change in the justice effect when performance info is provided
  # negative value = performance info reduces the punitive premium
  # the eliminated effect is the negative of this: -tau_interact > 0
  inquiry_interaction <- declare_inquiry(
    Justice_x_PerfInfo = tau_interact,
    label = "Justice x Perf Info Interaction"
  )

  # a: answer strategy -------------------------------------------------------
  # OLS with cluster-robust SEs at respondent level

  # (1) main effects model for AMCEs
  estimator_amce <- declare_estimator(
    Y ~ Z_justice + Z_threat + perf_positive + perf_negative,
    .method = lm_robust,
    clusters = respondent_id,
    se_type = "CR2",
    term = c("Z_justice", "Z_threat", "perf_positive", "perf_negative"),
    inquiry = c("AMCE_Justice", "AMCE_Threat",
                "AMCE_Perf_Positive", "AMCE_Perf_Negative"),
    label = "AMCEs (main effects)"
  )

  # (2) interaction model for eliminated effect (Acharya et al. 2018)
  # the coefficient on Z_justice:has_perf_info estimates how the justice effect
  # changes when performance info is provided; eliminated effect = -1 * this
  estimator_mechanism <- declare_estimator(
    Y ~ Z_justice * has_perf_info + Z_threat + perf_positive + perf_negative,
    .method = lm_robust,
    clusters = respondent_id,
    se_type = "CR2",
    term = "Z_justice:has_perf_info",
    inquiry = "Justice_x_PerfInfo",
    label = "Justice x Perf Info (interaction)"
  )

  # chain MIDA components
  model +
    assign_justice + assign_threat + assign_performance +
    measure_Y +
    inquiry_justice + inquiry_threat + inquiry_perf_pos + inquiry_perf_neg +
    inquiry_interaction +
    estimator_amce + estimator_mechanism
}

# instantiate with defaults
base_design <- declare_conjoint_design()

# run simulated experiments and compute diagnosands
diagnosis <- diagnose_design(base_design, sims = 500)

# display results
diagnosis$diagnosands_df |>
  select(
    Estimand    = inquiry,
    `True AMCE` = mean_estimand,
    Estimate    = mean_estimate,
    Bias        = bias,
    SD          = sd_estimate,
    RMSE        = rmse,
    Power       = power,
    Coverage    = coverage
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(
    caption = "Design diagnosis (500 simulations, N = 2,000, K = 5 tasks per respondent)",
    align   = c("l", rep("r", 7))
  )
```

**Interpretation:** The conjoint design with K = 5 tasks per respondent creates an effective analytical sample of N x K = 10,000 observations (for N = 2,000). Because attribute levels are independently randomized across tasks, each AMCE is estimated from the full set of observations. Cluster-robust standard errors account for the within-respondent correlation across the 5 tasks. The key advantage over the previous two-stage design is statistical efficiency: the performance dimension is no longer estimated within small sub-cells (N/8 per arm) but across the full analytical sample. Following @bansak2021, respondents can reliably evaluate up to 5--7 conjoint tasks without significant satisficing.

The **Justice x Perf Info interaction** row captures the eliminated effect logic from @acharya2018. A negative coefficient means that providing performance information *reduces* the punitive justice premium---evidence that implicit performance beliefs are part of the causal mechanism. The eliminated effect (ATE$_{\text{justice}}$ $-$ ACDE$_{\text{justice}}$) is the negative of this coefficient. In the simulation, the true interaction is $\tau_{\text{interact}}$ = -0.4, meaning that 0.4 scale points of the punitive premium (out of 0.8 total) are explained by implicit performance assumptions.

## Power by Sample Size

```{r}
#| label: fig-power-by-n
#| fig-cap: "Statistical power for each AMCE as a function of total sample size (N). K = 5 tasks per respondent. Dashed line marks the 80% power threshold."
#| cache: true

# vary sample size
n_vals <- c(500, 750, 1000, 1500, 2000, 2500)

designs_by_n <- setNames(
  lapply(n_vals, function(n) declare_conjoint_design(N = n)),
  paste0("N=", n_vals)
)

diagnoses_by_n <- diagnose_design(designs_by_n, sims = 500)

n_lookup <- tibble(
  design = paste0("N=", n_vals),
  N = n_vals
)

diagnoses_by_n$diagnosands_df |>
  left_join(n_lookup, by = "design") |>
  ggplot(aes(x = N, y = power, color = inquiry, group = inquiry)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  annotate("text", x = 500, y = 0.83, label = "80% threshold",
           hjust = 0, size = 3.2, color = "gray40") +
  scale_color_brewer(palette = "Set2", name = "Estimand") +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_x_continuous(labels = comma) +
  labs(x = "Total Respondents (N)", y = "Power") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```
